---
title: '`ggplot2`, pt. 2'
author: "Sarah Moore"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---
# Getting Started 

We are going to go through a workflow today to practice working with version control and saving from R directly. First, start a new project in your computer's local folder for this class. Call the R project _*anes_survey_work*_.

When you are in the project let's open an R script. Your working directory *should* be now set to whatever file you set as the path for your R project. So, it should be the file for this class. We can check this. So long as it's what we expected, we can just move along. 

```{r, eval = T, echo = F, results = 'hide'}

getwd()

```

What's next when we *VERY* first set out an R script? 

```{r, eval = T, echo = F, results = 'hide'}
# PACKAGES!!! 
library(tidyverse)
# install.packages("fauxnaif", dep = T)
# install.packages(c('ggthemes', "fauxnaif", "patchwork"))
library(ggthemes)
library(fauxnaif)
library(patchwork)

# and the data next!
anes2020 <- read_csv('https://raw.githubusercontent.com/sarah-moore/lousy-graphs/main/data/2020_ANES.csv')
glimpse(anes2020)
#8,280 observations over 1,771 columns 
```

Let's pick at max six variables that we want to work with just so that we don't have to get used to too much at once. 

How about these: 

a. **V200001**: Case ID 

b. **V201028**: Did the respondent vote for president

-9. Refused 

-1. Inapplicable 

1. Yes 

2. No
  
c. **V202024**: Has respondent gotten into a political argument in past 12 months

  -9. Refused 
  
  -7. Incomplete interview 
  
  -6. No post-election interview 
  
  1. Yes 
  
  2. No 

d. **V202025**: Has respondent joined a protest march, rally, or demonstration in past 12 months 

  -9. Refused 
  
  -7. Incomplete interview
  
  -6. No post-election interview 
  
  1. Yes
  
  2. No 
  
e. **V201014e**: Party registration in state of voter registration 

  -1. Inapplicable 
  
  0. No party registration 
  
  1. Party registration 
  
f. **V201115**: How hopeful respondent feels about how things are going in the country

  -9. Refused
  
  -8. Don't know 
  
  1. Not at all
  
  2. A little 
  
  3. Somewhat
  
  4. Very 
  
  5. Extremely 

g. **V201600**: Respondent sex 

-9. Refused 

1. Male 

2. Female 

Based on these variables we want to work with, subset them using `select()` and get rid of all the unwanted values. Give them new names on the way. 

```{r}
# subset 
anes2020 %>% 
  select(resp_id = V200001, 
         resp_pres_vote = V201028, 
         resp_pol_argument = V202024, 
         resp_part_protest = V202025, 
         resp_state_reg = V201014e, 
         resp_hopeful = V201115, 
         resp_sex = V201600) -> anes_sub
  
dim(anes_sub)

# figure out which values we need to rid ourselves of 
bad_values <- c(-1, -2, -5, -6, -7, -8, -9)

# use some fancy tricks and a new function to get the bad values out 
anes_sub %>%
  mutate(across(everything(), na_if_in, bad_values))-> anes_clean_sub

# then we can check if those values are indeed gone! 
any(anes_clean_sub[anes_clean_sub < 0])

any(bad_values %in% anes_clean_sub)
```

# What kinds of questions might motivate looking at these variables? 

# What kinds of variables are these??? 

- Out of the `geom` functions [here](https://ggplot2-book.org/individual-geoms.html), which make sense to use for univariate description?

- What about [bivariate or multivariate](https://ggplot2-book.org/collective-geoms.html)?

Using one of these ideas let's build from the bottom up. 

```{r, include= T, eval = F}
# let's start with the data. 
anes_clean_sub %>% 
  filter(!is.na(resp_pol_argument))%>% 
# now we map our x (and/or) y aesthetics, sometimes more if we want 
ggplot(aes(x = as_factor(resp_pol_argument))) + 
  geom_bar() + 
  ylim(0, 8280) + 
  scale_x_discrete(breaks = c("Yes" = 1, "No" = 2))
# let's set a geom_ layer 

# now run the code together!! and assign it a name 

# and now we'll save it 

ggsave("name_here.png", new_plot)
```

That was pretty basic, let's try a new variable(s) with a different `geom` function. PLUS add labels to the axes and let's change the colors 

```{r, include= T, eval = F}
# data 

# aes()

# geom_function() 

# labs()

# scale_brewer_manual()
               #-> new_plot2 

ggsave("name_here2.png", new_plot2)
```


Now we'll step away from these plots to check that they both made it to the specified folder. So long as they did, let's open up GitHub Desktop. If this folder you are working with is not yet a GH repo, let's make it one. If it is already, let's make a commit and a push to get these new images up on your GH repo. 

So long as the repo exists, you should see these images added in GH Desktop. Make a comment like "adding new plots" to the *Summary* and then commit the change. Then push it to the remote repo. 

Check your GH online repo to see if the push made it to the final destination. If all is good let's continue to move through some additional visualizations. 

Let's try a visualization that has some summary statistics in it. We'll have to think carefully about our *scale*. 

```{r}

anes_clean_sub$resp_pres_vote <- if_else(anes_clean_sub$resp_pres_vote== 2, 0, anes_clean_sub$resp_pres_vote)

anes_clean_sub %>% 
  filter(!is.na(resp_sex), !is.na(resp_state_reg)) %>% 
  ggplot(aes(y = resp_pres_vote, x =as_factor(resp_sex))) + 
  geom_point(stat = "summary", fun = "mean") + 
  ylim(0,1) + 
  facet_wrap(~resp_state_reg) 
# let's add labels here 

anes_clean_sub %>% 
  filter(!is.na(resp_sex), !is.na(resp_pol_argument)) %>% 
  ggplot(aes(y = resp_pres_vote, x =as_factor(resp_sex))) + 
  geom_point(stat = "summary", fun = "mean") + 
  ylim(0,1) + 
  facet_wrap(~resp_pol_argument) -> pol_argument_gg

anes_clean_sub %>% 
  filter(!is.na(resp_sex), !is.na(resp_part_protest)) %>% 
  ggplot(aes(y = resp_pres_vote, x =as_factor(resp_sex))) + 
  geom_point(stat = "summary", fun = "mean") + 
  ylim(0,1) + 
  facet_wrap(~resp_part_protest) -> pol_protest_gg
# let's add labels 

pol_argument_gg + pol_protest_gg -> argument_protest_gg 

ggsave("argument_protest.png", argument_protest_gg)
```
